% tbsrc/acceleration.tex — Acceleration and Noise: OSEM and Semi-convergence
% Status: written (first draft). OSEM (subsets, subset sensitivity, the
% speed/accuracy caveat); noise amplification and semi-convergence; early
% stopping; a brief regularization note. Matches osem/subset_models in
% src/reconstruction.jl.
\section{Acceleration and Noise: OSEM and Semi-convergence}
\label{sec:acceleration}

MLEM as written in Section~\ref{sec:mlem} is correct but slow and, run to
convergence, noisy. Two practical matters stand between \eqref{eq:ml-em} and a
usable image: each iteration touches every event, so many iterations are
expensive; and the maximum-likelihood image it converges to is itself noisy, so
converging fully is the wrong goal. Ordered subsets address the first,
early stopping the second. They are the difference between the algorithm and a
reconstruction recipe.

\subsection{Ordered-subsets EM (OSEM)}
\label{sec:acceleration:osem}

The cost of one MLEM update is dominated by the forward and back-projection over
the full event list. Ordered-Subsets EM~\cite{hudson1994} splits the events into
$M$ disjoint subsets and applies one EM update per subset, so a single pass
through the data --- one \emph{epoch} --- performs $M$ image updates instead of
one:
\begin{equation}
  \xv \;\leftarrow\; \frac{\xv}{\sensv^{(m)}}\;\odot\;
       \Aadj_{(m)}\!\Bigl(\code{mult}^{(m)}\odot
       \frac{\code{counts}^{(m)}}{\pred^{(m)}}\Bigr),
  \qquad m = 1,\dots,M,
  \label{eq:acc-osem}
\end{equation}
where subset $m$ carries its own events and its own \emph{subset sensitivity}
$\sensv^{(m)}=\Aadj_{(m)}(\code{mult}^{(m)})$ --- the back-projection of the
multiplicative factors over that subset's LORs, not the full set. This last point
is essential: each update is normalized by the detectability of the LORs it
actually uses, so the subset updates are on a common scale and one epoch advances
the image roughly as far as $M$ full MLEM iterations, an $M$-fold speed-up.

In the library, \code{subset\_models} splits the list into $M$ interleaved
subsets (subset $k$ is every $M$-th LOR from $k$, which keeps each subset's
angular coverage representative) and builds a \code{ListmodePoissonModel} per
subset, each with its own exact $\sensv^{(m)}$. \code{osem} then sweeps the
subsets once per epoch:
\begin{lstlisting}
models = subset_models(xs, xe, org, vs, n, nsub; counts = counts, mult = a)
rec    = osem(models, x0; nepochs = 10)   # 10 epochs ~ 10*nsub MLEM iterations
\end{lstlisting}

The acceleration is not free. The per-subset likelihood is not the full
likelihood, so OSEM is \emph{not} monotone and does not converge to a single
point: it settles into a limit cycle, drifting among $M$ nearby images rather
than landing on the ML solution. With many subsets and few counts per subset the
cycle is wide and the result biased. The usual compromise is a moderate subset
count that keeps each subset statistically representative; OSEM is then used to
reach a good image quickly, and a few plain MLEM iterations ($M=1$) can finish if
a stable fixed point is wanted.

\subsection{Noise and semi-convergence}
\label{sec:acceleration:semiconv}

The deeper issue is not speed. MLEM converges \emph{monotonically} to the
maximum-likelihood image (Section~\ref{sec:mlem}), but with Poisson data that
limit is a noisy image: the likelihood is maximized by an estimate that explains
not only the signal but the particular noise realization in the counts. Because
the update is multiplicative and unregularized, each iteration injects a little
more high-frequency detail, and late iterations fit noise. The likelihood keeps
improving while the image quality, after an early peak, gets worse --- the hallmark
of an ill-posed inverse problem.

This is \emph{semi-convergence}: the error against the true image falls, reaches
a minimum, then rises, even as $f$ decreases throughout
(Figure~\ref{fig:acc-semiconv} sketches the two curves). The reconstruction
passes through its best estimate at an intermediate iteration and degrades if
pushed further.
\begin{figure}[t]
  \centering
  \begin{tikzpicture}[x=1.1cm,y=0.9cm]
    % axes
    \draw[->] (0,0) -- (7.2,0) node[below right] {iteration $k$};
    \draw[->] (0,0) -- (0,4.3) node[above left] {};
    % negative log-likelihood: monotonically decreasing, flattening
    \draw[thick,blue,domain=0.15:7,samples=80,smooth]
      plot (\x, {3.6*exp(-0.55*\x)+0.35});
    \node[blue,right] at (3.6,1.05) {neg.\ log-likelihood $f$ (always $\downarrow$)};
    % error vs truth: U-shaped (semi-convergence)
    \draw[thick,red,domain=0.5:7,samples=90,smooth]
      plot (\x, {0.42*(\x-2.6)*(\x-2.6)+1.3});
    \node[red,right] at (4.4,3.5) {error vs truth};
    % minimum marker
    \draw[densely dashed,gray] (2.6,0) node[below,black]{\small stop} -- (2.6,1.3);
    \fill[red] (2.6,1.3) circle (1.6pt);
  \end{tikzpicture}
  \caption{Semi-convergence. The negative log-likelihood $f$ decreases at every
    iteration (MLEM is monotone), but the error against the true image is
    U-shaped: it reaches a minimum at an intermediate iteration and rises as
    later iterations fit the Poisson noise. The reconstruction is taken at the
    minimum --- \emph{early stopping}.}
  \label{fig:acc-semiconv}
\end{figure}

The practical remedy is \emph{early stopping}: take the image at a fixed, modest
iteration count rather than at convergence. The optimum depends on the count
level and the structure sought --- more counts and coarser features tolerate more
iterations --- so the iteration number is a reconstruction parameter, tuned per
study, not a constant. The worked examples stop at tens of iterations and report
the interior noise alongside the contrast to show the trade (Part~II): too few
iterations leave the image blurred and low-contrast, too many amplify the noise,
and a stopping point in between gives the best estimate. With OSEM the same
applies in epochs, scaled by the subset count.

\subsection{Beyond early stopping}
\label{sec:acceleration:reg}

Early stopping controls noise by halting before it dominates, which is implicit
regularization --- effective and parameter-light, but blunt: it applies one
global iteration budget to every region regardless of its local statistics. The
explicit alternative is to add a smoothness prior, reconstructing the maximum
\emph{a posteriori} image by minimizing $f(\xv)+\beta\,R(\xv)$ with $R$ penalizing
roughness and $\beta$ setting its strength. The preconditioned-gradient view of
Section~\ref{sec:mlem} makes this a small change: the same step
$\xv-(\xv/\sensv)\odot\nabla(\cdot)$ applies to the penalized objective, with the
prior's gradient added to $\nabla f$. RecoCrysp ships the unregularized MLEM/OSEM
and relies on early stopping; a penalty term is a natural extension of the same
update and is left for later work.
