% tbsrc/appendices.tex — Appendices.  STATUS: STUB (after \appendix in driver).
\section{Expectation--Maximization derivation of MLEM}
\label{app:em}

This appendix carries through the EM argument sketched in
Section~\ref{sec:mlem}, deriving the multiplicative update \eqref{eq:ml-em} from
the Poisson model. The forward model is $\pred_i = \code{mult}_i\,(\Aop\xv)_i +
\sct_i$ with $(\Aop\xv)_i=\sum_j A_{ij}x_j$, and the data are independent counts
$y_i\sim\mathrm{Poisson}(\pred_i)$ (Section~\ref{sec:statistics}). Here
$\code{mult}=\nrm\odot\att$, $\sct_i$ is the additive background
($\code{contamination}$), and $A_{ij}\ge0$ are the system-matrix entries.

\subsection*{Complete data}

EM introduces latent variables that, if observed, would make estimation trivial.
Split each LOR's counts by origin: let $N_{ij}$ be the number of counts on LOR
$i$ emitted from voxel $j$, and $N_{i0}$ the background counts. Decomposing the
Poisson mean $\pred_i$ over its independent contributions,
\begin{equation}
  N_{ij}\sim\mathrm{Poisson}\bigl(\code{mult}_i\,A_{ij}\,x_j\bigr),
  \qquad
  N_{i0}\sim\mathrm{Poisson}(\sct_i),
  \qquad
  y_i = \sum_j N_{ij} + N_{i0},
  \label{eq:app-em-latent}
\end{equation}
all independent; the sum reproduces $y_i\sim\mathrm{Poisson}(\pred_i)$ since
independent Poissons add. Treating $\{N_{ij}\}$ as observed, the complete-data
log-likelihood, dropping terms free of $\xv$ (the background and the
$\log N_{ij}!$ constants), is
\begin{align}
  \log L_{\mathrm c}(\xv)
  &= \sum_{i,j}\Bigl[N_{ij}\log\bigl(\code{mult}_i A_{ij} x_j\bigr)
        - \code{mult}_i A_{ij} x_j\Bigr] + \text{const} \notag\\
  &= \sum_j\Bigl[\Bigl(\textstyle\sum_i N_{ij}\Bigr)\log x_j
        - x_j\,\underbrace{\textstyle\sum_i \code{mult}_i A_{ij}}_{(\sensv)_j}\Bigr]
        + \text{const},
  \label{eq:app-em-complete}
\end{align}
where $(\sensv)_j=\sum_i \code{mult}_i A_{ij}=\Aadj(\code{mult})_j$ is the
sensitivity image. The activity decouples voxel by voxel: each $x_j$ sees only its
own total emission $\sum_i N_{ij}$ and its detectability $(\sensv)_j$.

\subsection*{E-step}

The latent counts are unknown, so EM replaces them by their conditional
expectation given the data and the current estimate $\xv^k$. Conditioned on the
total $y_i$, the counts on LOR $i$ are distributed among the voxels and the
background by a multinomial with probabilities proportional to their Poisson
means, hence
\begin{equation}
  \mathbb E\bigl[N_{ij}\mid y_i,\xv^k\bigr]
  = y_i\,\frac{\code{mult}_i A_{ij} x_j^k}{\pred_i^k},
  \qquad
  \pred_i^k=\code{mult}_i\,(\Aop\xv^k)_i+\sct_i .
  \label{eq:app-em-estep}
\end{equation}
Substituting into \eqref{eq:app-em-complete} gives the $Q$-function
$Q(\xv\mid\xv^k)=\sum_j\bigl[\bar M_j^k\log x_j - x_j(\sensv)_j\bigr]+\text{const}$,
with the expected voxel emission
\begin{equation}
  \bar M_j^k
  = \sum_i \mathbb E\bigl[N_{ij}\mid y_i,\xv^k\bigr]
  = x_j^k\sum_i A_{ij}\,\frac{\code{mult}_i\,y_i}{\pred_i^k}
  = x_j^k\,\Aadj\!\Bigl(\code{mult}\odot\frac{y}{\pred^k}\Bigr)_j .
  \label{eq:app-em-mbar}
\end{equation}

\subsection*{M-step}

Maximizing the concave $Q$ over $x_j\ge0$ is a one-dimensional problem per voxel.
Setting $\partial Q/\partial x_j = \bar M_j^k/x_j - (\sensv)_j = 0$ yields
$x_j^{k+1}=\bar M_j^k/(\sensv)_j$, i.e.\ with \eqref{eq:app-em-mbar}
\begin{equation}
  x_j^{k+1}
  = \frac{x_j^k}{(\sensv)_j}\,
    \Aadj\!\Bigl(\code{mult}\odot\frac{y}{\pred^k}\Bigr)_j ,
  \label{eq:app-em-update}
\end{equation}
which is exactly the multiplicative MLEM update \eqref{eq:ml-em} (with $y=\code{counts}$).
Each step is non-negative when $\xv^k$ is, and a voxel at zero stays at zero
(Section~\ref{sec:mlem}).

\subsection*{Why the likelihood climbs}

EM increases the observed-data likelihood at every step. Writing $\ell(\xv)$ for
the observed log-likelihood, the general EM identity
$\ell(\xv)=Q(\xv\mid\xv^k)+H(\xv\mid\xv^k)$ holds, where $H$ is the conditional
entropy of the latent variables; by Gibbs' inequality
$H(\xv\mid\xv^k)\ge H(\xv^k\mid\xv^k)$ for all $\xv$. Since the M-step makes
$Q(\xv^{k+1}\mid\xv^k)\ge Q(\xv^k\mid\xv^k)$, adding the two inequalities gives
$\ell(\xv^{k+1})\ge\ell(\xv^k)$: the negative log-likelihood
\code{neg\_log\_likelihood} is monotonically non-increasing, as
Section~\ref{sec:mlem} asserts.

\section{De Pierro's surrogate for quadratic priors}
\label{app:depierro}

This appendix derives the closed-form penalized update
\eqref{eq:reg-depierro} from the MAP objective
$\Phi(\xv)=f(\xv)+\beta R(\xv)$ for a quadratic prior, following De
Pierro~\cite{depierro1995}. The likelihood part $f$ is handled by the EM surrogate
of Appendix~\ref{app:em}: at the current iterate $\xv^k$ it is majorized by the
separable
\begin{equation}
  Q_f(\xv\mid\xv^k) = \sum_j\bigl[\,t_j\log x_j - (\sensv)_j\,x_j\,\bigr] + \text{const},
  \qquad
  t_j = x_j^k\,\Aadj\!\Bigl(\code{mult}\odot\tfrac{\code{counts}}{\pred^k}\Bigr)_j,
  \label{eq:app-dp-Qf}
\end{equation}
with $t_j$ the EM numerator \eqref{eq:reg-numerator}. The penalty must be made
separable too.

\subsection*{Separable surrogate for a quadratic penalty}

Take the smoothness prior $R(\xv)=\tfrac12\sum_j\sum_{k\in\mathcal N_j}(x_j-x_k)^2$
(unit weights; the intensity prior is the easier diagonal case). The coupling is
through the difference $x_j-x_k$, which De Pierro rewrites about the current
iterate as
\begin{equation}
  x_j-x_k
  = \tfrac12\bigl[\,2x_j-(x_j^k+x_k^k)\,\bigr] - \tfrac12\bigl[\,2x_k-(x_j^k+x_k^k)\,\bigr],
\end{equation}
an average of two terms each depending on a single voxel. Convexity of the square,
$\bigl(\tfrac12 a-\tfrac12 b\bigr)^2\le\tfrac12 a^2+\tfrac12 b^2$, then gives the
separable upper bound
\begin{equation}
  (x_j-x_k)^2 \;\le\; \tfrac12\bigl[2x_j-(x_j^k+x_k^k)\bigr]^2
                    + \tfrac12\bigl[2x_k-(x_j^k+x_k^k)\bigr]^2,
  \label{eq:app-dp-ineq}
\end{equation}
with equality at $\xv=\xv^k$ (both bracketed terms become $x_j^k-x_k^k$ and the
average is exact). Summing \eqref{eq:app-dp-ineq} over the symmetric neighbourhood
defines a surrogate $\widetilde R(\xv\mid\xv^k)\ge R(\xv)$ that touches $R$ at
$\xv^k$ and is a sum of one-voxel quadratics: collecting the terms in $x_j$,
\begin{equation}
  \widetilde R(\xv\mid\xv^k)
  = \sum_j\Bigl[\,\tfrac12\gamma_j\,x_j^2 - c_j\,x_j\,\Bigr] + \text{const},
  \qquad
  \gamma_j = 2\beta\,w_j,
  \quad
  c_j = \beta\,(w_j x_j^k + m_j),
  \label{eq:app-dp-Rtilde}
\end{equation}
where $w_j=\abs{\mathcal N_j}$ is the in-bounds neighbour count and
$m_j=\sum_{k\in\mathcal N_j}x_k^k$ the neighbour sum at $\xv^k$ (the $\beta$
reinstated). The intensity prior $\tfrac{\beta}{2}\norm{\xv-\bm z}^2$ is already
separable, contributing $\gamma_j=\beta$, $c_j=\beta z_j$ directly.

\subsection*{M-step}

Maximizing the total surrogate $Q_f-\widetilde R$ over $x_j\ge0$ is again one
problem per voxel. Setting the derivative to zero,
\begin{equation}
  \frac{t_j}{x_j} - (\sensv)_j - \gamma_j x_j + c_j = 0
  \;\;\Longrightarrow\;\;
  \gamma_j\,x_j^2 + \underbrace{\bigl[(\sensv)_j - c_j\bigr]}_{B_j}\,x_j - t_j = 0,
\end{equation}
a quadratic with $G_j=\gamma_j$ and $B_j=(\sensv)_j-c_j$. Substituting
\eqref{eq:app-dp-Rtilde} gives exactly the coefficients of
Table~\ref{tab:reg-priors}: $B_j=(\sensv)_j-\beta(w_jx_j^k+m_j)$, $G_j=2\beta w_j$
for the smoothness prior and $B_j=(\sensv)_j-\beta z_j$, $G_j=\beta$ for the
intensity prior. The non-negative root,
\begin{equation}
  x_j^{k+1} = \frac{-B_j+\sqrt{B_j^2+4G_jt_j}}{2G_j}
            = \frac{2t_j}{\sqrt{B_j^2+4G_jt_j}+B_j},
\end{equation}
is \eqref{eq:reg-depierro}; the right-hand form is used in code because it stays
finite as $G_j\to0$, where it returns $t_j/B_j=t_j/(\sensv)_j$ --- the MLEM update.
Because the surrogate majorizes $\Phi$ and is tight at $\xv^k$, decreasing it
decreases $\Phi$: the iteration is monotone.

\section{Notation}
\label{app:notation}
\textit{(to be written: full glossary of symbols.)}

\section{RecoCrysp API quick reference}
\label{app:api}
\textit{(to be written: the functions and model fields used in the examples.)}
