% tbsrc/geometry.tex — Scanner Geometry and the Line of Response
% Status: written (first draft). Covers both detector models (pixelated and
% continuous/monolithic) and the listmode vs sinogram data forms.
\section{Scanner Geometry and the Line of Response}
\label{sec:geometry}

Everything the reconstruction needs to know about the scanner reduces to one
thing: the set of lines of response and, for each, how many coincidences it
recorded. This section fixes how a LOR is represented, how two very different
detector technologies --- pixelated crystal rings and monolithic blocks ---
both produce that representation, and how the resulting data are stored.

\subsection{The endpoint model}
\label{sec:geometry:endpoint}

A line of response is defined entirely by its two ends. We write them as
world-space points $\bm{x}_s,\bm{x}_e\in\RR^3$, in millimetres, in a fixed
right-handed frame whose $z$ axis is the scanner (axial) axis and whose
$(x,y)$ plane is transverse. A scanner that records $M$ coincidences therefore
hands the reconstruction two $3\times M$ matrices of endpoints, and nothing
else about the detector enters the projector
(Section~\ref{sec:projection}):
\begin{equation}
  \texttt{xstart} = \bigl[\bm{x}_s^{(1)}\;\cdots\;\bm{x}_s^{(M)}\bigr],
  \qquad
  \texttt{xend} = \bigl[\bm{x}_e^{(1)}\;\cdots\;\bm{x}_e^{(M)}\bigr].
  \label{eq:geom-endpoints}
\end{equation}

This \emph{endpoint-driven} design is the central modelling choice of
\code{RecoCrysp}. The projector integrates a thin line between whatever two
points it is given (Section~\ref{sec:projection:joseph}); it has no notion of
crystals, blocks, or detector pitch. Two consequences follow. First, the
\emph{geometry} of the scanner --- where coincidences can occur --- and its
\emph{resolution} --- how precisely each is localized --- are completely
separate concerns: resolution is modelled on its own
(Section~\ref{sec:resolution}), never baked into $\Aop$. Second, any detector
technology is supported as soon as it can deliver endpoint pairs. The next two
subsections are exactly that: two ways to populate
\eqref{eq:geom-endpoints}.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[scale=1.1]
    \draw[thick] (0,0) circle (2.4);
    \draw[dashed,gray] (0,0) circle (1.25);
    \node[gray] at (0,-0.78) {\small FOV};
    \coordinate (s) at (128:2.4);
    \coordinate (e) at (-37:2.4);
    \draw[blue,very thick] (s) -- (e);
    \fill (s) circle (1.6pt) node[above left] {$\bm{x}_s$};
    \fill (e) circle (1.6pt) node[above right] {$\bm{x}_e$};
    \fill (0,0) circle (1pt);
    \draw[->,gray] (0,0) -- (0.8,0) node[below right] {\small $x$};
    \draw[->,gray] (0,0) -- (0,0.8) node[above left] {\small $y$};
  \end{tikzpicture}
  \caption{Transverse section of a cylindrical scanner. A coincidence defines a
    line of response between two detector points $\bm{x}_s,\bm{x}_e$; the
    annihilation lies somewhere along it (non-TOF). The dashed circle is the
    transverse field of view (Section~\ref{sec:grid}).}
  \label{fig:geom-lor}
\end{figure}

\subsection{Pixelated (discrete-crystal) detectors}
\label{sec:geometry:pixelated}

The conventional PET detector is a ring, or stack of rings, of small
scintillator crystals. Each crystal has a fixed centre, and a coincidence
between crystals $a$ and $b$ is assigned the LOR joining their two centres.
The endpoints are therefore drawn from a \emph{discrete, finite} set, and the
transverse spacing of the LORs is quantized by the crystal pitch; the
achievable resolution is correspondingly limited to roughly half a crystal
width (Section~\ref{sec:resolution}).

\code{RecoCrysp} builds such a scanner as a regular polygon of crystals,
following \code{parallelproj}~\cite{schramm2024}:
\begin{lstlisting}
scanner = RegularPolygonPETScannerGeometry(;
    radius = 100.0, num_sides = 48, num_lor_endpoints_per_side = 1,
    lor_spacing = 4.0, ring_positions = -12.0:4.0:12.0, symmetry_axis = 3)
\end{lstlisting}
The crystal world coordinates are precomputed; a \emph{LOR descriptor} then
enumerates the detector pairs and lays them out in sinogram order
(Section~\ref{sec:geometry:data}), yielding the endpoint matrices
\eqref{eq:geom-endpoints}.

\subsection{Monolithic (continuous) detectors}
\label{sec:geometry:continuous}

A monolithic detector replaces the crystal array with a large, undivided
scintillator block (in CRYSP~\cite{crysp2026}, $50\times50\,\mathrm{mm}^2$
faces, a few radiation lengths deep). There are no crystals to read out; instead
a neural network estimates the three-dimensional interaction position
$(x,y,z)$ from the scintillation-light pattern. The decisive consequence is
that the LOR endpoint can fall \emph{anywhere} on the detector surface, not on a
discrete grid, and that the spatial resolution is set by the
detector-plus-network --- about $3.5\,\mathrm{mm}$ FWHM for CRYSP --- and is
\emph{decoupled from the $50\,\mathrm{mm}$ block size}.

The faithful model is therefore a \emph{continuous} cylindrical surface, not a
ring of points:
\begin{lstlisting}
scanner = ContinuousPET(diameter = 774.0, afov = 1024.0)  # CRYSP, mm
xstart, xend = sample_lors(scanner, n_events; rng = rng)
\end{lstlisting}
\code{sample\_lors} draws LOR endpoints continuously over the cylinder
(here the CRYSP bore diameter $77.4\,\mathrm{cm}$ and axial field of view
$102.4\,\mathrm{cm}$). The finite detector resolution is then applied as a
separate, detector-level step (Section~\ref{sec:resolution}); it is not part of
the geometry.

\begin{remark}
It is tempting to model a monolithic scanner by snapping endpoints to the
centres of its blocks, treating each block as a giant ``crystal''. This is
wrong: it would cap the resolution at half a block ($\sim25\,\mathrm{mm}$) and
discard exactly the continuous positioning that the detector provides. The
block size is a packaging detail; the resolution that matters
($3.5\,\mathrm{mm}$) is a property of the position estimate, and is injected
separately. Sampling endpoints continuously and modelling resolution on its own
is what keeps the two decoupled.
\end{remark}

\subsection{Listmode and sinogram data}
\label{sec:geometry:data}

Whatever the detector, the recorded coincidences can be stored in two ways. In
\emph{listmode}, every event is one record carrying its two endpoints,
$(x_1,y_1,z_1,x_2,y_2,z_2)$; the data are just the columns of
\eqref{eq:geom-endpoints}, optionally with a per-event weight. In a
\emph{sinogram}, the discrete LORs are binned and one stores the count on each
bin, indexed by radial offset, view (angle), and axial plane. The pixelated
descriptor produces its LORs already in a chosen sinogram order.

Listmode is the natural form for a continuous detector --- there is no discrete
bin to assign a continuously-positioned event to --- and it is the
representation this tutorial uses throughout. Either way, reconstruction sees
the same two ingredients: a list of LORs (the endpoints) and a per-LOR count
(Section~\ref{sec:statistics}). With the geometry fixed, the next section places
the image on a voxel grid in the same world frame.
