% tbsrc/mlem.tex — Iterative Reconstruction: MLEM
% Status: written (first draft). EM motivation -> multiplicative update; MLEM as
% a preconditioned gradient step x - (x/sens) grad f (the form RecoCrysp uses);
% the dual role of the sensitivity image; non-negativity, monotonicity, support.
\section{Iterative Reconstruction: MLEM}
\label{sec:mlem}

Section~\ref{sec:statistics} reduced reconstruction to minimizing the convex
objective $f$ of \eqref{eq:st-listmode}. A plain gradient descent on $f$ would
need a step size and could step $\xv$ negative, neither of which suits an
activity image. The Maximum-Likelihood Expectation-Maximization (MLEM) algorithm
of Shepp and Vardi~\cite{shepp1982}, extended to the present model by Lange and
Carson~\cite{lange1984}, avoids both: a parameter-free multiplicative update that
keeps $\xv\ge\bm0$ automatically and decreases $f$ at every step. This section
states the update, shows it is the preconditioned gradient step the library
implements, and reads off its properties.

\subsection{The EM update}
\label{sec:mlem:em}

EM treats reconstruction as a missing-data problem. The counts $\code{counts}$
are observed, but \emph{which voxel} each detected coincidence originated from is
not. If those voxel-of-origin assignments were known, the maximum-likelihood
activity would just be their tally divided by detectability --- a trivial
estimate. EM iterates that idea: given the current image, the E-step computes the
\emph{expected} contribution of each voxel to each LOR's counts; the M-step
re-estimates the image from those expectations. Carried through (Appendix~A), the
two steps collapse into one multiplicative update,
\begin{equation}
  \xv_{k+1} \;=\; \frac{\xv_k}{\sensv}\;\odot\;
                  \Aadj\!\Bigl(\code{mult}\odot\frac{\code{counts}}{\pred_k}\Bigr),
  \qquad
  \pred_k = \code{mult}\odot(\Aop\xv_k) + \code{contamination},
  \label{eq:ml-em}
\end{equation}
with $\odot$ and the division taken elementwise and
$\sensv=\Aadj(\code{mult})$. Each iteration is one forward projection (to form
the prediction $\pred_k$), one ratio, and one back-projection --- the two Joseph
operations of Section~\ref{sec:projection}, nothing more. The correction factor
$\Aadj(\code{mult}\odot\code{counts}/\pred_k)$ compares measured to predicted
counts and back-projects the mismatch: where the model under-predicts, the ratio
exceeds one and the voxel is scaled up; at the fixed point every ratio averages
to one and the image stops changing.

\subsection{MLEM as a preconditioned gradient step}
\label{sec:mlem:precond}

The library does not code \eqref{eq:ml-em} as written; it codes the equivalent
\emph{preconditioned gradient} form, which exposes what the update is doing.
Using the gradient $\nabla f = \sensv - \Aadj(\code{mult}\odot\code{counts}/\pred)$
from \eqref{eq:st-grad}, scale it by the diagonal preconditioner
$\bm D_k=\diag(\xv_k/\sensv)$ and take one descent step of unit length:
\begin{equation}
  \xv_{k+1} \;=\; \xv_k \;-\; \frac{\xv_k}{\sensv}\odot\nabla f(\xv_k)
            \;=\; \xv_k - \frac{\xv_k}{\sensv}\odot\Bigl(\sensv -
                  \Aadj(\tfrac{\code{mult}\,\code{counts}}{\pred_k})\Bigr)
            \;=\; \frac{\xv_k}{\sensv}\odot\Aadj\!\Bigl(
                  \frac{\code{mult}\,\code{counts}}{\pred_k}\Bigr).
  \label{eq:ml-precond}
\end{equation}
The $\sensv$ in the first parenthesis cancels the $\xv_k/\sensv$ scaling, leaving
exactly \eqref{eq:ml-em}: \emph{the multiplicative MLEM update is gradient descent
with step size one and the diagonal preconditioner $\xv_k/\sensv$}. This is
\code{em\_update} in \code{reconstruction.jl} --- written subtractively as
$\xv_k-(\xv_k/\sensv)\odot\nabla f$ --- and it explains the absence of a step
size: the preconditioner sets the scale of every voxel's step, large where the
image is bright and detectability low, small where it is dim or well-sampled.

\subsection{The sensitivity image, twice over}
\label{sec:mlem:sens}

The sensitivity image $\sensv=\Aadj(\code{mult})=\Aadj(\nrm\odot\att)$ does two
jobs in \eqref{eq:ml-precond}, and conflating them is a common source of
confusion. As the \emph{normalizer} it is the gradient's first term ---
$\inner{\sensv}{\xv}$ in the objective --- accounting for each voxel's total
detection probability, so that a voxel seen by few LORs is not mistaken for one
with little activity. As the \emph{preconditioner} it is the $1/\sensv$ scaling
of the step. Both call for the same vector, the back-projection of the
multiplicative factors over \emph{all} geometric LORs, computed once by
\code{sensitivity\_image} before iterating (Section~\ref{sec:datamodel}). Because
it is decoupled from the event list, the data term in \eqref{eq:ml-em} projects
only over events while $\sensv$ carries the all-LOR normalization --- the listmode
economy of Section~\ref{sec:statistics} made concrete.

\subsection{Properties}
\label{sec:mlem:properties}

Four properties follow from the multiplicative form, and they are why MLEM is the
default.

\begin{itemize}
  \item \emph{Non-negativity.} Started from a positive image, every factor in
        \eqref{eq:ml-em} is non-negative, so $\xv_k\ge\bm0$ for all $k$ with no
        projection step --- the constraint is built in. The library writes the
        update subtractively for speed and clamps at zero
        (\code{max.(..., 0)}), since Float32 cancellation in
        \eqref{eq:ml-precond} can leave rounding-level negatives.
  \item \emph{Monotonicity.} EM guarantees the likelihood never worsens:
        $f(\xv_{k+1})\le f(\xv_k)$ at every iteration~\cite{lange1984}, so
        \code{neg\_log\_likelihood} decreases monotonically --- a direct
        convergence check, with no tuned step size to destabilize it.
  \item \emph{Support preservation.} A voxel that reaches zero stays zero: the
        multiplicative update cannot resurrect it. Voxels outside the field of
        view have $\sensv=0$ and are left untouched (the preconditioner masks
        them), so the reconstruction lives on the support the geometry actually
        sees.
  \item \emph{Fixed point.} At convergence $\nabla f=\bm0$, i.e.\
        $\Aadj(\code{mult}\odot\code{counts}/\pred)=\sensv$: the
        prediction-weighted back-projected data balance the sensitivity
        everywhere. This is the maximum-likelihood image for the model
        \eqref{eq:dm-full}.
\end{itemize}

\noindent
Monotone convergence to the maximum-likelihood image is not the same as
convergence to a \emph{good} image: with noisy data the ML solution is itself
noisy, and MLEM approaches it through a useful intermediate before degrading.
That behaviour --- semi-convergence --- and the ordered-subsets acceleration that
makes the iterations affordable are the subject of Section~\ref{sec:acceleration}.
