% tbsrc/statistics.tex — Statistical Model and the Likelihood
% Status: written (first draft). Poisson counting; the Poisson log-likelihood;
% the <sens,x> term and the listmode/sinogram equivalence; the negative-log-
% likelihood objective and its gradient, matching src/reconstruction.jl.
\section{Statistical Model and the Likelihood}
\label{sec:statistics}

The data model \eqref{eq:dm-full} gives the \emph{mean} counts $\pred$ on each
LOR. What an acquisition actually records is a random draw around that mean.
Reconstruction is therefore not the inversion of a linear system --- there is no
$\xv$ with $\Aop\xv$ equal to the noisy data --- but a statistical
\emph{estimation}: find the image whose predicted means make the observed counts
most probable. This section writes down the probability of the data given the
image, turns it into an objective to minimize, and computes its gradient. The
next section turns that gradient into the MLEM update.

\subsection{Poisson counting}
\label{sec:statistics:poisson}

Annihilations occur independently at a rate fixed by the activity, and each
detected coincidence is a rare event among enormously many photon histories.
Counting such events over a fixed acquisition makes the number recorded on LOR
$i$ a Poisson variable with mean $\pred_i$,
\begin{equation}
  y_i \sim \mathrm{Poisson}(\pred_i),
  \qquad
  P(y_i \mid \pred_i) = \frac{\pred_i^{\,y_i}\,e^{-\pred_i}}{y_i!},
  \label{eq:st-poisson}
\end{equation}
and counts on distinct LORs are independent. Two consequences set the whole
reconstruction strategy. First, the variance equals the mean, so a LOR with few
expected counts is noisy in \emph{relative} terms --- the fractional uncertainty
is $1/\sqrt{\pred_i}$ --- which is why low-count regions reconstruct poorly and
why the estimator must respect counting statistics rather than treat every LOR
equally. Second, because the LORs are independent, the probability of the whole
dataset is the product over LORs, and a sum over LORs once we take logarithms.

\subsection{The Poisson log-likelihood}
\label{sec:statistics:loglik}

The likelihood of the image is the probability of the observed counts as a
function of $\xv$ (through $\pred=\pred(\xv)$). Multiplying \eqref{eq:st-poisson}
over all LORs and taking the negative logarithm gives, dropping the
$\log y_i!$ terms that do not depend on $\xv$,
\begin{equation}
  f(\xv) \;=\; \sum_i \Bigl[\,\pred_i - y_i \log \pred_i\,\Bigr],
  \qquad
  \pred_i = \code{mult}_i\,(\Aop\xv)_i + \code{contamination}_i .
  \label{eq:st-nll}
\end{equation}
This $f$ is the negative Poisson log-likelihood; the maximum-likelihood image is
its minimizer. It is convex in $\xv$ (the first term is linear, $-\log$ is
convex, and $\pred$ is linear in $\xv$), so a single global minimum is sought ---
the difficulty is not local minima but noise, treated in
Section~\ref{sec:acceleration}.

\subsection{The sensitivity term and the listmode form}
\label{sec:statistics:listmode}

The two sums in \eqref{eq:st-nll} are not summed over the same thing, and seeing
why is what makes listmode reconstruction efficient. Take the first sum and
substitute $\pred_i$, dropping the additive background (constant in $\xv$):
\begin{equation}
  \sum_i \pred_i \;=\; \sum_i \code{mult}_i\,(\Aop\xv)_i + \text{const}
  \;=\; \inner{\code{mult}}{\Aop\xv} + \text{const}
  \;=\; \inner{\Aadj(\code{mult})}{\xv} + \text{const}
  \;=\; \inner{\sensv}{\xv} + \text{const},
  \label{eq:st-sens}
\end{equation}
using the adjoint $\inner{\bm w}{\Aop\xv}=\inner{\Aadj\bm w}{\xv}$
(Section~\ref{sec:projection}) and the sensitivity image
$\sensv=\Aadj(\code{mult})$ of \eqref{eq:dm-pred}. This term runs over
\emph{every geometric LOR the scanner can record}: it is the total detectability
of the image, and it is precomputed once as the back-projection $\sensv$,
independent of any counts.

The second sum, $-\sum_i y_i \log \pred_i$, has a factor $y_i$ in every term, so
only LORs with $y_i>0$ contribute --- the \emph{detected events}. This is the
listmode economy: rather than predict and store a count for all $\sim\!10^{12}$
possible LORs, evaluate the data term over the event list alone, while the
all-LOR first term collapses into the single precomputed vector $\sensv$.
Writing $\code{counts}$ for the per-event counts (one per raw event, or an
integer multiplicity for compressed listmode), the objective becomes
\begin{equation}
  f(\xv) \;=\; \inner{\sensv}{\xv}
            \;-\; \sum_{\text{events}} \code{counts}\,\log \pred ,
  \qquad
  \sensv=\Aadj(\code{mult}),
  \label{eq:st-listmode}
\end{equation}
which is exactly \code{neg\_log\_likelihood} in \code{reconstruction.jl}. The
sinogram and listmode likelihoods are the same function \eqref{eq:st-nll}
rearranged: the sinogram form bins counts onto a fixed LOR grid and sums over
bins; the listmode form sums over events and folds all undetected LORs into
$\sensv$. They agree term by term, and the listmode form is the natural one for a
continuous scanner whose events do not fall on any grid
(Section~\ref{sec:geometry}).

\subsection{The gradient}
\label{sec:statistics:gradient}

Minimizing $f$ needs its gradient. The first term in \eqref{eq:st-listmode} is
linear, with $\nabla\inner{\sensv}{\xv}=\sensv$. For the data term, the chain
rule through $\pred_i=\code{mult}_i(\Aop\xv)_i+\code{contamination}_i$ gives
$\partial \pred_i/\partial x_j = \code{mult}_i\,A_{ij}$, so
\begin{equation}
  \nabla f(\xv) \;=\; \sensv \;-\; \Aadj\!\Bigl(\code{mult}\odot
                     \frac{\code{counts}}{\pred}\Bigr),
  \label{eq:st-grad}
\end{equation}
the back-projection of the per-LOR ratio $\code{mult}\odot\code{counts}/\pred$
subtracted from the sensitivity --- the \code{\_gradient} routine, with the
convention $0/0=0$ on unmeasured LORs. The structure is telling: at the minimum
$\nabla f=0$ means $\Aadj(\code{mult}\odot\code{counts}/\pred)=\sensv$, i.e.\ the
back-projected, prediction-weighted data balance the sensitivity everywhere. The
gradient mixes a forward projection (to form $\pred$) and a back-projection (to
spread the ratio into image space), the two Joseph operations the library
provides. Section~\ref{sec:mlem} turns \eqref{eq:st-grad} into a multiplicative
update that preserves non-negativity and needs no step-size tuning.
